#Solution for *Using principle components as covariates* {.solution}
We start by loading all relevant data.

```{r loadData}
geno <- readr::read_tsv(file("/data/genotypes/genotypes.tab.gz"))
expr <- readr::read_tsv(file("/data/monocytes/expression/ifn_expression.tab.gz"))
```

## Size of dataset

```{r dataSize}
dim(expr)
dim(geno)
```

Note that these files have samples in columns and variables in rows. 
So the data consists of `r ncol(expr) - 1` samples with measurements for `r nrow(expr)` 
gene expression probes and `r nrow(geno)` SNPs.
  
## Computing principle components
R provides the function `prcomp` for this task. Like most standard R functions
it expects data to be laid out with variables in columns and samples in rows.
We therefore have to transpose the data, compute and extract the principle
components (stored in the `x` element of the return value).

```{r prcomp}
pca <- prcomp(t(expr[-1]), center=TRUE, scale = TRUE)
pc <- pca$x
```
Plotting the variances for the first 20 PCs is then straightforward.

```{r pcPlot}
plot(pca, npcs=10)
```

Since the data were scaled prior to the PCA the total variance is the same as
the number of probes. The variance accounted for by each component is available
through the `sdev` field of the `prcomp` return value.

```{r varProp}
sum(pca$sdev[1:10]^2)/nrow(expr)
```

## Fitting a model with PC covariates
To make our life a bit easier we collect all the relevant data into a single *data.frame*.

```{r preparePCfit}
data <- data.frame(probe=unlist(subset(expr, Probe=="3710685")[-1]), 
		rs4077515=unlist(subset(geno, id=="rs4077515")[-1]), pc[,1:10])
```

Now we fit the model including the PCs:

```{r fitPC}
pcFit <- lm(probe ~ ., data=data)
summary(pcFit)
```

For comparison we also fit the simple model:

```{r fitSimple}
simpleFit <- lm(probe ~ rs4077515, data=data)
summary(simpleFit)
```

## Visualising SNP effect on gene expression
As in the previous set of exercises we plot the gene expression with the
effect of the non-genetic covariates removed.

```{r rs4077515Plot}
library(ggplot2)
corrected <- data$probe - rowSums(coef(pcFit)[-(1:2)]*data[, 3:12])
corrected <- data.frame(expression=corrected, genotype=factor(data$rs4077515))
ggplot(corrected, aes(genotype, expression)) +
		geom_jitter(colour="darkgrey", position=position_jitter(width=0.25)) +
		geom_boxplot(outlier.size=0, alpha=0.6, fill="grey") + theme_bw()
```
